% =============================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================

We presented a graph neural network framework for satellite-based solar
irradiance forecasting that addresses the situational-awareness needs of
solar-integrated power system operators in tropical climates. The key
contributions are: (i) a weighted $k$-NN GraphSAGE-LSTM architecture that
explicitly models spatial cloud structure from GOES-16 patches; (ii) a
rigorous comparative evaluation against a ResNet-LSTM baseline, a flat-MLP
spatial-ablation baseline, and persistence at two Colombian sites with
contrasting climatic regimes; and (iii) Split Conformal Prediction as a
distribution-free post-hoc UQ layer that transforms point forecasts into
coverage-guaranteed prediction intervals usable in risk-aware operational
decisions.

\textbf{GraphSAGE-LSTM} achieves the best skill among the deep-learning models
at the 6-hour horizon at \emph{both} sites (El~Paso: $\skillday = 0.636$ vs.\
$0.614$ for ResNet-LSTM; Uniandes: $0.417$ vs.\ $0.368$), with lower seed
variance throughout, confirming that distance-weighted neighbourhood
aggregation on the pixel graph provides a more stable and expressive spatial
representation than fixed convolutional filters.

At the \textbf{tropical mountain site} (Uniandes), however, the absolute skill
of every model is markedly lower than at El~Paso even at the most favourable
horizon ($\skillday \leq 0.42$ vs.\ $0.64$ at 6h), revealing that the strong
diurnal autocorrelation of the Bogot{\'a} clear-sky cycle and the locally-driven
nature of orographic convection limit the marginal benefit of satellite
spatial features at this site. This finding has a direct operational
implication: model selection and skill expectations should be climate-aware,
and forecasting pipelines should always be benchmarked and calibrated against
the local persistence reference rather than assumed to transfer across sites.

At the \textbf{1-hour horizon}, no model consistently beats persistence at El~Paso,
consistent with the low persistence error ($204.4$~W/m²) in a predominantly
clear-sky regime. Modest positive skill ($\approx 0.15$) is achieved at Uniandes,
where frequent cloud passages create exploitable structure.

\textbf{Future work} follows a staged uncertainty-quantification roadmap
toward fully probabilistic, operationally-deployable solar forecasts.
Split Conformal Prediction has been implemented as Stage~1;
Conformalized Quantile Regression \cite{romano2019} will yield adaptive-width
intervals that contract during predictable clear-sky conditions and expand
during uncertain cloud events (Stage~2), enabling tighter reserve scheduling.
Stage~3 will achieve full Bayesian decomposition of forecast uncertainty:
Gaussian variance heads will capture aleatoric uncertainty (irreducible cloud
stochasticity), while Stochastic Gradient Langevin Dynamics (SGLD,
\cite{welling2011}) will sample the posterior weight distribution to provide
epistemic uncertainty estimates \cite{kendall2017}, completing the transition
from point forecasts to risk-stratified probabilistic outputs for
solar-integrated power system operation.
Additional planned work includes the FlatMLP spatial-ablation analysis,
multi-step joint-horizon outputs, and extension to further Latin American sites
to build a broader tropical solar forecasting benchmark.
