% =============================================================
\section{Discussion}
\label{sec:discussion}
% =============================================================

\subsection{Why does GraphSAGE outperform ResNet at longer horizons?}

At the 6-hour horizon for El~Paso, GraphSAGE-LSTM achieves a 2.2-percentage-point
skill improvement over ResNet-LSTM ($0.636$ vs.\ $0.614$). We hypothesise two
mechanisms:

\textbf{Irregular spatial structure.} At longer lead times, the relevant cloud
signal migrates from the centre of the patch (where it directly causes GHI
reduction) toward the periphery, where incoming cloud advection is reflected in
the satellite channels. CNN filters, being translation-equivariant, apply the same
spatial template across all positions; GraphSAGE, by contrast, learns separate
aggregation rules that can weight distant pixels differently depending on their
graph neighbourhood. This flexibility may be especially valuable for detecting
cloud-edge gradients that precede GHI events.

\textbf{Reduced redundancy.} The ResNet applies multiple stacked convolutions
that effectively redundantly process the spatial information across layers; the
GraphSAGE message-passing is applied once per hop and is less prone to
over-smoothing at the patch scale ($P = 16$). The lower seed variance of
GraphSAGE ($\pm 0.013$ vs.\ $\pm 0.023$ at El~Paso 6h) is consistent with a
more stable loss landscape.

\subsection{Why does SARIMAX remain competitive at Uniandes 6h?}

The finding that SARIMAX ($\skillday = 0.435$) outperforms GraphSAGE-LSTM ($0.417$)
at the 6-hour horizon for Uniandes is counterintuitive given the richer information
available to the DL models. We attribute it to the \emph{predictability structure}
of the Bogot{\'a} climate:

\textbf{Diurnal regularity.} At $4.6^{\circ}\text{N}$, the solar declination varies little
across the year ($\pm 4.6^{\circ}$), producing a highly repeatable clear-sky envelope.
Cloud formation in Bogot{\'a} is predominantly triggered by orographic lifting in
the early afternoon, creating a systematic diurnal pattern that the seasonal MA
and AR terms of SARIMAX($\cdot$)(1,1,1)$_{24}$ can capture efficiently.

\textbf{Satellite signal saturation.} At 6 hours, the satellite patch at $t$
may have low predictive relevance for the ground GHI at $t+6\text{h}$: the cloud
system that will arrive has not yet entered the patch. SARIMAX effectively uses
the historical autocorrelation of $k_t$ (which encodes the regular afternoon cloud
cycle) as its predictive feature, a signal that does not degrade with forecast
horizon in the same way that satellite spatial features do.

This result suggests that \emph{site-specific climatological structure should inform
model selection}: in climates with strong, regular diurnal cloud patterns,
statistical baselines may remain competitive at multi-hour horizons even when
high-resolution satellite data are available.

\subsection{Why is the 1-hour horizon intractable at El~Paso?}

The negative skill at 1-hour El~Paso (all DL models, $\skillday \approx -0.17$
to $-0.20$) can be explained by the \emph{low persistence error}: at 204.4~W/m²,
persistence is already very accurate because the predominantly clear-sky regime
means irradiance rarely drops by more than a small fraction over 10 minutes.
The DL models, trained to minimise MSE over the full distribution (including
the occasional cloud event), learn a slightly smoothed version of the target
trajectory that incurs higher error on the flat clear-sky majority.

A potential remedy is a loss function that differentially weights cloud events
(e.g., weighted MSE by clear-sky departure $|k_t - 1|$), or dedicated cloud-event
detection as a preprocessing step.

\subsection{Practical implications}

\begin{itemize}
  \item For \textbf{El~Paso-type sites} (high irradiance, moderate variability),
        GraphSAGE-LSTM with Optuna-tuned hyperparameters is the recommended model
        at 3h and 6h horizons, with a 60+\% reduction in RMSE over persistence.
        At 1h, the system should fall back to persistence or a light-weight
        nowcasting model.

  \item For \textbf{Uniandes-type sites} (tropical mountain, high convective
        variability), SARIMAX on the clear-sky index provides a surprisingly strong
        baseline at 6h and should be included in any benchmarking suite. DL models
        add modest value but require careful hyperparameter tuning to outperform it.

  \item The high seed variance of ResNet-LSTM at El~Paso 3h ($\pm 0.090$) suggests
        that multiple retraining runs should be averaged for production deployment,
        or that GraphSAGE-LSTM should be preferred for its higher stability.
\end{itemize}

\subsection{Uncertainty quantification: from point forecasts to calibrated intervals}
\label{ssec:uq_discussion}

Grid operators require probabilistic outputs — prediction intervals that are
well-calibrated across conditions of varying cloud cover — not just point
estimates. Quantifying and ultimately decomposing forecast uncertainty is the
central long-term goal of this work.

\textbf{Stage 1: Split Conformal Prediction (implemented).}
Because all models produce only point forecasts, we apply Split CP
\cite{vovk2005,angelopoulos2023} as a distribution-free post-hoc layer.
The calibrated interval $[\hat{y} \pm \hat{q}]$ (Section~\ref{ssec:uq})
carries a finite-sample marginal coverage guarantee regardless of the model
residual distribution, which is particularly valuable for tropical sites where
the error distribution is heavy-tailed due to cloud passages.
A key limitation is that $\hat{q}$ is globally constant: the interval does not
narrow during predictable clear-sky mornings nor widen during uncertain afternoon
convection at Uniandes.

\textbf{Stage 2: Conformalized Quantile Regression (planned).}
CQR \cite{romano2019} replaces the symmetric residual score with a
quantile-regression score that adapts the interval width to local difficulty.
This is expected to reduce average interval width substantially at El~Paso
(predominantly clear-sky) while maintaining coverage.

\textbf{Stage 3: Aleatoric/epistemic decomposition (planned).}
Gaussian variance heads (NLL loss) and MC~Dropout will provide separate
estimates of irreducible measurement noise (aleatoric) and model-knowledge
uncertainty (epistemic). Under the formulation of \cite{kendall2017}, SGLD
\cite{welling2011} can further sample the posterior weight distribution for a
fully Bayesian treatment.

\subsection{Implications for CPES resilience}

The results presented here connect directly to the resilience of solar-integrated
cyber-physical energy systems.

\textbf{Situational awareness under cloud-induced uncertainty.}
GraphSAGE-LSTM, by learning spatially-aware representations of incoming cloud
systems from satellite imagery, provides operators with earlier detection of
impending irradiance ramps --- the primary source of generation-side uncertainty
in PV-heavy CPES. The 63.6\% RMSE reduction over persistence at 6h for El~Paso
translates to significantly tighter dispatch intervals and reduced reserve
requirements.

\textbf{Calibrated prediction intervals for risk-aware dispatch.}
The Split CP layer converts any point forecast into an interval with a
finite-sample coverage guarantee. For a CPES operator, this means that reserve
scheduling can be tied to an explicit coverage target (e.g., 90\% intervals)
rather than to ad hoc safety margins. The global constant width $2\hat{q}$ is
an acknowledged limitation, but even symmetric intervals provide stronger
risk-management guarantees than unquantified point forecasts.

\textbf{Climate-adaptive model selection.}
The finding that SARIMAX outperforms deep learning at Uniandes~6h has a direct
operational implication: CPES operators in tropical mountain locations should
benchmark statistical baselines before deploying resource-intensive DL pipelines.
A hybrid portfolio strategy --- statistical baseline for climatologically
predictable periods, spatial DL during high-variability windows --- could
optimise the resilience-cost tradeoff.

\subsection{Limitations}

\textbf{Geographic scope.} Both sites are located in Colombia and represent
two tropical climate sub-types. Generalisation to other tropical or subtropical
regions requires additional validation.

\textbf{Flat MLP ablation in progress.} The FlatMLP experiments are currently
completing their Optuna runs (24/24 configurations being evaluated). Final
results will be incorporated to quantify what fraction of deep-learning skill
arises from spatial encoding versus temporal dynamics alone.

\textbf{Single-step output.} Current models produce a single-step forecast at
horizon $H$. Simultaneous multi-step output (forecasts for all three horizons
from a single forward pass) could improve consistency and reduce inference cost
in real-time CPES applications.
