\section{Introduction}
\label{sec:intro}

\IEEEPARstart{M}{odern} power grids are increasingly characterised as
cyber-physical energy systems (CPES) \cite{deng2017} --- tightly coupled
infrastructures in which physical generation, transmission, and distribution
assets are monitored and controlled through digital communication layers.
The integration of variable renewable energy (VRE), particularly solar
photovoltaic (PV), introduces a new source of operational uncertainty that
challenges the \emph{resilience} of CPES: the ability to anticipate,
withstand, and recover from disruptive events \cite{bhusal2020}.
Unlike cyber attacks or extreme weather damage --- which are discrete and
locatable --- the uncertainty arising from cloud-induced irradiance variability
is continuous, stochastic, and correlated across spatially distributed
generation assets.
Accurate, uncertainty-aware short-term solar forecasting is therefore a
foundational \emph{AI-enabled resilience mechanism} for CPES operators,
providing the situational awareness needed to schedule reserves, trigger demand
response, and prevent grid imbalances before they cascade.

Solar photovoltaic (PV) generation has grown rapidly across Latin America, with
Colombia reaching over 1~GW of installed capacity and targeting 9~GW by 2030
under its National Energy Plan \cite{upme2023}. Reliable short-term forecasts of
global horizontal irradiance (GHI) are a prerequisite for efficient dispatch of
PV-integrated power systems: they enable reserve scheduling, reduce imbalance
penalties, and support real-time grid management. In tropical countries such as
Colombia, the forecasting challenge is compounded by the high frequency and spatial
heterogeneity of convective cloud systems, which can reduce surface irradiance by
more than 80\% within minutes and are poorly captured by coarse-resolution numerical
weather prediction (NWP) models.

Satellite-based nowcasting methods bypass NWP limitations by inferring cloud
motion and optical depth directly from geostationary imagery. Instruments such as
the GOES-16 Advanced Baseline Imager (ABI) provide multi-spectral observations
every 10~minutes at sub-kilometre resolution, making them well-suited for
sub-hourly to hourly irradiance estimation and forecasting. The dominant paradigm
couples a spatial feature extractor (applied to a patch of satellite pixels
centred on the target site) with a temporal model that exploits the recent
history of observed GHI or satellite state \cite{zhang2015,reno2016}.

Deep learning has substantially advanced this paradigm. Convolutional neural
networks (CNNs) applied to satellite patches learn translation-invariant spatial
patterns associated with cloud regimes, and when combined with recurrent models
such as LSTMs they achieve state-of-the-art performance on benchmark datasets
\cite{alessandrini2015,huertas2018,paletta2021}. More recently, graph neural networks
(GNNs) have emerged as an alternative spatial encoder: by representing pixels as
nodes in a graph and learning neighbourhood aggregation rules, GNNs relax the
translation-invariance assumption of CNNs and can capture irregular spatial
dependencies introduced by non-uniform cloud structures \cite{hamilton2017,ji2022}.
However, systematic comparisons of GNN-based and CNN-based encoders for GHI
forecasting remain scarce, and none, to the best of our knowledge, address the
tropical Andean climate.

\subsection*{Contributions}

This paper makes the following contributions:

\begin{enumerate}
  \item We propose \textbf{GraphSAGE-LSTM}, a model that treats a GOES-16 satellite
        patch as a $k$-nearest-neighbour weighted pixel graph, applies GraphSAGE
        neighbourhood aggregation \cite{hamilton2017} with inverse-distance edge
        weights, and passes the resulting sequence of graph embeddings through an
        LSTM to produce a point forecast of GHI at horizons of 1, 3, and 6~hours.

  \item We conduct a \textbf{rigorous empirical comparison} across two climatologically
        distinct Colombian sites, three forecast horizons, and four random seeds,
        including Bayesian hyperparameter optimisation via Optuna \cite{akiba2019}
        for GraphSAGE-LSTM, ResNet-LSTM, and a flat MLP spatial-ablation baseline.

  \item We show that GraphSAGE-LSTM \textbf{consistently outperforms} ResNet-LSTM
        at the 6-hour horizon for the low-elevation high-irradiance site
        ($\skillday = 0.636$ vs.\ $0.614$), while the advantage diminishes for
        the high-altitude, high-variability site.

  \item We identify a \textbf{regime-dependence} of statistical vs.\ deep-learning
        methods: SARIMAX on the clear-sky index outperforms deep-learning models
        at 6~hours for the tropical mountain site (Bogot{\'a}, $\skillday = 0.435$
        vs.\ $0.417$), where the regular diurnal cycle dominates over
        cloud-driven spatial variability.

  \item We implement \textbf{Split Conformal Prediction} \cite{vovk2005,angelopoulos2023}
        as a post-hoc UQ layer applicable to any trained backbone.
        The calibration procedure produces marginal-coverage-valid prediction
        intervals without distributional assumptions, supporting risk-aware
        dispatch decisions in CPES. This constitutes Stage~1 of a staged UQ
        roadmap: Split~CP (this work) $\to$ Conformalized Quantile Regression
        \cite{romano2019} for adaptive-width intervals $\to$ Bayesian
        uncertainty decomposition \cite{kendall2017} for aleatoric/epistemic
        separation.

  \item We establish a \textbf{reproducible GOES-16 benchmark for tropical sites},
        including data manifests, patch extraction, model code, Optuna studies,
        and evaluation scripts, addressing the scarcity of published benchmarks
        for Latin American and equatorial CPES contexts.
\end{enumerate}

\subsection*{Paper organisation}

Section~\ref{sec:related} reviews related work. Section~\ref{sec:methodology}
describes the datasets, model architectures, training protocol, and evaluation
metrics. Section~\ref{sec:results} presents quantitative results.
Section~\ref{sec:discussion} analyses key findings. Section~\ref{sec:conclusion}
concludes and outlines future work.
