\section{Methodology}
\label{sec:methodology}
\subsection{Study sites}

Experiments are conducted at two meteorological stations in Colombia that represent
climatologically contrasting regimes (Table~\ref{tab:sites}).

\begin{table}[t]
  \centering
  \caption{Study site metadata.}
  \label{tab:sites}
  \begin{tabular}{lcc}
    \toprule
    & \textbf{El Paso (César)} & \textbf{Uniandes (Bogotá)} \\
    \midrule
    Latitude  & $9.737°\text{N}$ & $4.602°\text{N}$ \\
    Longitude & $73.695°\text{W}$ & $74.066°\text{W}$ \\
    Elevation & 50~m a.s.l.      & 2\,600~m a.s.l.  \\
    Climate   & Tropical semi-arid & Equatorial mountain \\
    Avg.\ GHI & ${\sim}$5.5~kWh/m²/day & ${\sim}$4.2~kWh/m²/day \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{El Paso} is located in the lowland Caribbean region of C{\'e}sar department,
characterised by high annual irradiance, predominantly clear-sky conditions, and
dry-season cloud cover driven by the Inter-Tropical Convergence Zone (ITCZ).

\textbf{Universidad de los Andes} (referred to as Uniandes) is an instrumented
rooftop station in the Colombian Andean city of Bogot{\'a}. At 2\,600~m above sea
level, it experiences the equatorial bimodal rainfall regime with intense convective
cloud formation typically peaking in the early afternoon, producing sharp intraday
GHI drops.

\subsection{Data}

\subsubsection{Ground measurements}

GHI is measured at 10-minute resolution and stored in time-zone-aware (UTC) Parquet
files. Missing values account for less than 2\% of the record at both sites and are
excluded during dataset construction.

\subsubsection{Satellite data}

We use the GOES-16 ABI Level-2 Multi-Band Cloud and Moisture Imagery Product
(ABI-L2-MCMIPF), which provides all 16 ABI spectral channels at a unified
10-minute cadence and ${\sim}2$~km spatial resolution in the visible bands.
For each observation timestamp, a $16 \times 16$ pixel patch ($P = 16$) centred
on the target site is extracted, yielding a tensor of shape $(C, P, P) = (16, 16, 16)$
in units consistent with each band's native reflectance or brightness temperature.

\subsubsection{Dataset construction and temporal split}

For the deep-learning models (ResNet-LSTM, GraphSAGE-LSTM, FlatMLP), forecast
samples are constructed as pairs $(X, y)$ where the input
$X \in \R^{L \times C \times P \times P}$ is a sequence of $L = 24$ satellite
patch frames covering the most recent 4 hours of observation, and the target
$y \in \R$ is the GHI at the site $H$ steps ahead ($H \in \{6, 18, 36\}$ steps,
corresponding to 1, 3, and 6 hours at 10-minute resolution).
A pre-computed satellite manifest validates coverage for each timestamp and
filters out samples where any of the $L$ history frames is missing.
\textbf{SARIMAX} does not use satellite data or the patch manifest; it operates
directly on the 1D ground GHI series resampled to hourly resolution
(Section~\ref{ssec:sarima}).

To prevent data leakage, train, validation, and test sets are defined by
non-overlapping calendar blocks (Table~\ref{tab:splits}); the same boundaries
apply to both the satellite-based DL pipeline and the SARIMAX ground series.

\begin{table}[t]
  \centering
  \caption{Temporal data splits.}
  \label{tab:splits}
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{llccc}
    \toprule
    \textbf{Site} & \textbf{Split} & \textbf{Start} & \textbf{End} & \textbf{Samples (approx.)} \\
    \midrule
    \multirow{3}{*}{El Paso}
      & Train & 2022-03-01 & 2023-06-30 & 69\,984 \\
      & Val   & 2023-07-01 & 2023-10-31 & 17\,568 \\
      & Test  & 2023-11-01 & 2024-03-07 & 18\,288 \\
    \midrule
    \multirow{3}{*}{Uniandes}
      & Train & 2023-09-01 & 2024-09-30 & 56\,850 \\
      & Val   & 2024-10-01 & 2024-12-31 & 13\,104 \\
      & Test  & 2025-01-01 & 2025-03-28 & 12\,384 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Models}

\subsubsection{Persistence baseline}

The persistence model assumes irradiance is unchanged over the forecast horizon:
$\hat{y}(t+H) = \text{GHI}(t)$. It is optimal under no-change dynamics and
serves as the standard benchmark for short-term solar forecasting.

\subsubsection{SARIMAX on the clear-sky index}
\label{ssec:sarima}

Raw GHI is non-stationary due to the deterministic diurnal envelope. We first
compute the clear-sky index $k_t$:
\begin{equation}
  k_t = \frac{\text{GHI}_t}{\ghics_t}, \quad
  k_t = 0 \text{ if } \ghics_t < 10\,\text{W/m}^2
  \label{eq:csi}
\end{equation}
where $\ghics$ is the Ineichen clear-sky model \cite{ineichen2002} evaluated with
\texttt{pvlib} \cite{holmgren2018}. The series $k_t$ lies in $[0, 1.5]$ and is
more stationary than raw GHI.

Because the 10-minute resolution would require an intractable seasonal period of
$m = 144$, we resample $k_t$ to hourly resolution ($m = 24$). Order selection
via ADF/KPSS tests, ACF/PACF analysis, and AIC grid search \cite{hyndman2018}
yields the model:
\begin{equation}
  \text{SARIMAX}(2,1,2)(1,1,1)_{24}
  \label{eq:sarima}
\end{equation}

The fitted model forecasts $k_t$ out-of-sample; the final GHI prediction is
recovered as $\hat{\text{GHI}}(t) = \hat{k}(t)\cdot\ghics(t)$.

\subsubsection{ResNet-LSTM}
\label{ssec:resnet}

The ResNet-LSTM encoder-decoder processes the patch sequence $X$ as follows.
A lightweight ResNet \cite{he2016} (three residual stages with stride-2 downsampling,
followed by global average pooling) encodes each frame $(C, P, P)$ into a
$d_e$-dimensional embedding $\mathbf{e}_\ell \in \R^{d_e}$. The sequence
$(\mathbf{e}_1, \ldots, \mathbf{e}_L)$ is then processed by a single- or two-layer
LSTM \cite{hochreiter1997} whose final hidden state is projected to the scalar
output $\hat{y}$.

The ResNet encoder has three hyperparameters: base channel width
(\texttt{base}~$\in\{16, 24, 32, 48\}$), embedding dimension
(\texttt{emb\_dim}~$\in\{64, 128, 192\}$), and head dropout. The total parameter
count ranges from approximately 50k to 500k depending on the configuration.

\subsubsection{GraphSAGE-LSTM}
\label{ssec:gsage}

Instead of convolutional layers, we model the spatial patch as a weighted graph
$G = (V, E, \mathbf{w})$ where each pixel is a node ($|V| = P^2 = 256$).
For each node $v$ we connect its $k$ nearest neighbours in the 2D pixel grid
by Euclidean distance, and assign each edge an inverse-distance weight:
\begin{equation}
  w_{vu} = \frac{1}{d(v,u)}, \quad (v, u) \in E.
  \label{eq:edge_weight}
\end{equation}
This $k$-NN graph is more flexible than a fixed 8-connected grid: it can represent
irregular spatial dependencies and the optimal $k$ is treated as a hyperparameter
(Section~\ref{ssec:optuna}). Node features at step $\ell$ are the $C = 16$
channel values at that pixel.

GraphSAGE \cite{hamilton2017} updates each node's representation by aggregating
its neighbourhood. With edge weights, the mean aggregation becomes a
distance-weighted sum:
\begin{equation}
  \mathbf{h}_v^{(k)} = \sigma\!\left(
    \mathbf{W}^{(k)} \left[ \mathbf{h}_v^{(k-1)} \Vert
    \frac{\sum_{u \in \mathcal{N}(v)} w_{vu}\,\mathbf{h}_u^{(k-1)}}
         {\sum_{u \in \mathcal{N}(v)} w_{vu}} \right]
  \right)
  \label{eq:sage}
\end{equation}
where $\mathcal{N}(v)$ denotes the $k$-NN neighbours of $v$, $\Vert$ is
concatenation, and $\sigma$ is ReLU. After $K \in \{2,3,4\}$ layers, a mean
readout over all nodes yields the graph-level embedding
$\mathbf{g}_\ell \in \R^{d_g}$. The sequence
$(\mathbf{g}_1, \ldots, \mathbf{g}_L)$ is then processed by a single- or
two-layer LSTM whose final hidden state is projected to the scalar output $\hat{y}$.

The edge index and weights are fixed buffers registered with the model and
replicated for each batch element, so standard mini-batch DataLoaders and the
existing training loop are reused without modification.

\subsubsection{Flat MLP baseline}
\label{ssec:mlp}

As an ablation of spatial structure, we include a \textbf{FlatMLP} baseline that
applies global average pooling over the $P \times P$ spatial dimensions of each
frame, yielding a flattened feature vector of size $L \times C = 24 \times 16 = 384$.
This vector is passed through a multi-layer perceptron with \texttt{LayerNorm} and
\texttt{ReLU} activations. By discarding all spatial arrangement information,
the FlatMLP quantifies how much of the forecasting skill of the spatial models
(ResNet-LSTM and GraphSAGE-LSTM) arises from spatial encoding per se, versus
simple temporal statistics of the patch mean. The MLP follows the same Optuna
HPO protocol as the other deep-learning models (50 trials, 4 seeds, 2 sites,
3 horizons). Experiments are currently being finalised; results will be
incorporated in the version submitted for review.

\subsection{Hyperparameter optimisation}
\label{ssec:optuna}

For the three deep-learning models (ResNet-LSTM, GraphSAGE-LSTM, and FlatMLP)
we use Optuna \cite{akiba2019} with the TPE sampler and Median pruner.
The study budget is 50 trials per configuration
(site $\times$ horizon $\times$ seed) with $n_\text{jobs}=2$; an extended
run with 100 trials is being evaluated to confirm convergence.
SARIMAX does not use Optuna: its model order is selected via statistical tests
and information-criterion grid search as described in Section~\ref{ssec:sarima}.

The ResNet-LSTM and GraphSAGE-LSTM search spaces (Table~\ref{tab:hpspace})
include \texttt{n\_lstm\_layers}~$\in\{1,2\}$ to allow stacked recurrent
processing, and a categorical L1 regularisation coefficient
(\texttt{l1\_reg}) to penalise large weights in high-capacity configurations;
both are new relative to the initial baseline study.
For \textbf{ResNet-LSTM} specifically, the embedding dimension
(\texttt{emb\_dim}) and temporal hidden size (\texttt{hidden\_t}) are extended
to 256 to increase capacity.
For \textbf{GraphSAGE-LSTM}, the number of nearest neighbours
($k \in \{4, 8, 12, 16\}$) used to build the weighted pixel graph is
additionally tunable, allowing the study to determine the optimal spatial
connectivity for each site-horizon combination.
For \textbf{FlatMLP}, the study searches over MLP depth, hidden dimension,
dropout, learning rate, and weight decay (50 trials).

\begin{table}[t]
  \centering
  \caption{Optuna hyperparameter search spaces.}
  \label{tab:hpspace}
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{lll}
    \toprule
    \textbf{Hyperparameter} & \textbf{ResNet-LSTM} & \textbf{GraphSAGE-LSTM} \\
    \midrule
    \texttt{base} / \texttt{hidden\_g}  & $\{16,24,32,48\}$          & $\{64,96,128,192,256\}$      \\
    \texttt{emb\_dim} / \texttt{n\_sage\_layers} & $\{64,128,192,256\}$ & $\{2,3,4\}$            \\
    \texttt{hidden\_t} (LSTM)  & $\{64,128,192,256\}$       & $\{64,96,128,192,256\}$      \\
    \texttt{n\_lstm\_layers}   & $\{1,2\}$                  & $\{1,2\}$                    \\
    \texttt{k\_neighbors}      & ---                        & $\{4,8,12,16\}$              \\
    \texttt{l1\_reg}           & $\{0,10^{-5},10^{-4},10^{-3}\}$ & $\{0,10^{-5},10^{-4},10^{-3}\}$ \\
    \texttt{dropout}           & $\{0,0.1,0.2,0.3\}$        & $\{0,0.1,0.2\}$              \\
    \texttt{lr}                & $[3\text{e-}4, 3\text{e-}3]$ & $[5\text{e-}5, 2\text{e-}3]$ \\
    \texttt{weight\_decay}     & $[10^{-6}, 10^{-3}]$       & $[10^{-6}, 10^{-3}]$         \\
    \texttt{batch\_size}       & $\{16,32,64\}$             & $\{8,16,32\}$                \\
    \bottomrule
  \end{tabular}
\end{table}

The Optuna study identifies the best trial on validation $\rmseday$; the winning
configuration is then retrained for up to 30~epochs with patience~8 and evaluated
on the held-out test set.

\subsection{Training protocol}

All deep-learning models are trained with the Adam optimiser minimising the MSE
loss on normalised targets ($y_\text{norm} = (y - \bar{y}) / s_y$, with $\bar{y}$
and $s_y$ estimated from the training set). Training uses early stopping with
patience~6 (trials) or~8 (final retraining). Automatic Mixed Precision (AMP) is
enabled when a GPU is available. Each experiment is replicated with four seeds
$\{42, 1, 7, 13\}$ to report mean~$\pm$~std statistics.

\subsection{Uncertainty quantification: Split Conformal Prediction}
\label{ssec:uq}

All point-forecast models are used as the backbone for a three-stage uncertainty
quantification (UQ) pipeline whose long-term goal is to decompose total forecast
uncertainty into its \emph{aleatoric} component (irreducible cloud stochasticity)
and its \emph{epistemic} component (uncertainty due to limited data and model
capacity).

\textbf{Stage~1 --- Split Conformal Prediction (implemented, \cite{vovk2005,angelopoulos2023}).}
We apply Inductive Conformal Prediction as a distribution-free, finite-sample
calibration method.  A non-conformity score is computed over the validation set:
\begin{equation}
  s_i = |y_i - \hat{y}_i| \quad [\text{W/m}^2].
  \label{eq:score}
\end{equation}
The calibrated conformal quantile at miscoverage level $\alpha$ is
\begin{equation}
  \hat{q} = \mathrm{Quantile}\!\left(\{s_i\}_{i=1}^{n_\text{cal}},\;
    \frac{\lceil(n_\text{cal}+1)(1-\alpha)\rceil}{n_\text{cal}}\right),
  \label{eq:qhat}
\end{equation}
and the prediction interval for a new test point is
\begin{equation}
  \mathcal{C}_{1-\alpha}(x) = \bigl[\hat{y}(t+H)-\hat{q},\;\hat{y}(t+H)+\hat{q}\bigr].
  \label{eq:interval}
\end{equation}
Under exchangeability, this guarantees marginal coverage
$\mathbb{P}(Y \in \mathcal{C}_{1-\alpha}) \geq 1-\alpha$ exactly in finite samples
\cite{angelopoulos2023}.  Because $\hat{q}$ is a global constant, the interval
has fixed width $2\hat{q}$ and does not adapt to local prediction difficulty.
We evaluate empirical coverage at $\alpha \in \{0.05, 0.10, 0.20\}$ (intervals
of 95\%, 90\%, and 80\%) reporting both overall and daytime-only metrics.

\textbf{Stage~2 --- Conformalized Quantile Regression (planned, \cite{romano2019}).}
To obtain intervals whose width adapts to local uncertainty, we will train
quantile-regression heads on the DL encoders and apply the CQR correction.
This is expected to reduce unnecessary interval width during low-variability
periods (clear-sky mornings) without sacrificing coverage.

\textbf{Stage~3 --- Bayesian uncertainty decomposition (planned,
\cite{kendall2017,welling2011}).}
Gaussian variance heads (NLL loss) will model aleatoric uncertainty, while
Stochastic Gradient Langevin Dynamics (SGLD, \cite{welling2011}) will sample
the posterior weight distribution to quantify epistemic uncertainty.
SGLD injects calibrated Langevin noise into gradient updates, enabling
approximate Bayesian inference without the computational overhead of full
MCMC: running the chain for $T$ epochs and collecting snapshots every
$\Delta$ epochs yields $T/\Delta$ approximate posterior samples, whose
ensemble mean serves as the point forecast and whose standard deviation
quantifies model-knowledge uncertainty.
Together, the aleatoric head and SGLD ensemble will decompose total forecast
uncertainty into its irreducible (cloud stochasticity) and reducible
(model-knowledge) components \cite{kendall2017}, supporting risk-stratified
dispatch decisions in CPES.

\subsection{Evaluation metrics}

Following standard practice in solar forecasting \cite{coimbra2013}, we report:

\textbf{Daytime RMSE} ($\rmseday$, W/m²): RMSE computed only over test samples
with observed $\text{GHI} \geq 20$~W/m², excluding night-time samples where all
models trivially predict zero.

\textbf{Daytime skill score} ($\skillday$): normalised improvement over persistence,
\begin{equation}
  \skillday = 1 - \frac{\rmseday(\text{model})}{\rmseday(\text{persistence})}
  \label{eq:skill}
\end{equation}
A score of 0 indicates parity with persistence; positive values indicate improvement.
We also report $\text{MAE}_\text{day}$ for completeness.