% =============================================================
\section{Discussion}
\label{sec:discussion}
% =============================================================

\subsection{Why does GraphSAGE outperform ResNet at the 6-hour horizon?}

At the 6-hour horizon for El~Paso, GraphSAGE-LSTM achieves a 2.2-percentage-point
skill improvement over ResNet-LSTM ($0.636$ vs.\ $0.614$). We hypothesise two
mechanisms:

\textbf{Irregular spatial structure (hypothesis).} We hypothesise that, at
longer lead times, the relevant cloud signal migrates from the centre of the
patch (where it directly causes GHI reduction) toward the periphery, where
incoming cloud advection would be reflected in the satellite channels. CNN
filters, being translation-equivariant, apply the same spatial template
across all positions; GraphSAGE, by contrast, learns separate aggregation
rules that can weight distant pixels differently depending on their graph
neighbourhood. This flexibility may be especially valuable for detecting
cloud-edge gradients that precede GHI events, though we do not directly
verify the hypothesised centre-to-periphery migration here (Section~\ref{sec:discussion}, Limitations).

\textbf{Reduced redundancy (hypothesis).} We further hypothesise that the
ResNet's multiple stacked convolutions redundantly process spatial
information across layers, while GraphSAGE's message-passing, applied once
per hop, is less prone to over-smoothing at the patch scale ($P = 16$); we
do not directly measure representation redundancy or over-smoothing here.
The lower seed variance of GraphSAGE ($\pm 0.013$ vs.\ $\pm 0.023$ at
El~Paso 6h) is consistent with a more stable loss landscape, though as noted
in Section~\ref{ssec:optuna} this variance also reflects independently
re-run hyperparameter search outcomes across seeds, not training
stochasticity alone.

\subsection{Why is forecasting skill lower at Uniandes than at El~Paso?}

Across all deep-learning models and horizons, absolute skill is markedly lower
at the Uniandes mountain site than at El~Paso (e.g., GraphSAGE-LSTM at 6h:
$\skillday = 0.417$ vs.\ $0.636$). We attribute this gap to two related
properties of the Bogot{\'a} climate:

\textbf{Diurnal regularity narrows the room for improvement.}
At $4.6^{\circ}\text{N}$, the solar declination varies little across the year
($\pm 4.6^{\circ}$), producing a highly repeatable clear-sky envelope. Cloud
formation in Bogot{\'a} is predominantly triggered by orographic lifting in
the early afternoon, generating a strong, regular diurnal cycle. Much of the
predictable structure in the Uniandes GHI series is therefore already encoded
in its own recent history and time-of-day, leaving comparatively less
\emph{additional} information for a 4-hour satellite patch sequence to
contribute --- which compresses the achievable skill ceiling for every model,
spatially-aware or not.

\textbf{Satellite signal saturation at longer horizons (hypothesis).} At 6
hours, the satellite patch observed at $t$ may carry little predictive
information about the ground GHI at $t + 6\,\text{h}$: we hypothesise that
the cloud system that will eventually affect the site has often not yet
entered the $16\times16$-pixel field of view, particularly at a site where
convection forms locally through orographic lifting rather than advecting in
from a stable upstream direction. If so, this would limit how much spatial
encoding --- regardless of architecture --- can contribute at the longest
horizon; we do not directly measure cloud advection speed against the patch
footprint here.

This result suggests that \emph{site-specific climatological structure should
inform both model design and expectation-setting}: in climates with strong,
regular diurnal cloud patterns and locally-forming convection, the marginal
benefit of spatial deep learning over simpler temporal signals is smaller, and
reported skill scores should always be interpreted relative to the local
persistence baseline rather than compared at face value across sites.

\subsection{Why is the 1-hour horizon intractable at El~Paso?}

The negative skill at 1-hour El~Paso (all DL models, $\skillday \approx -0.17$
to $-0.20$) can be explained by the \emph{low persistence error}: at 204.4~W/m²,
persistence is already very accurate because the predominantly clear-sky regime
means irradiance rarely drops by more than a small fraction over 10 minutes.
We hypothesise that the DL models, trained to minimise MSE over the full
distribution (including the occasional cloud event), learn a slightly
smoothed version of the target trajectory that incurs higher error on the
flat clear-sky majority; we do not directly inspect residuals here to
confirm this smoothing.

A potential remedy is a loss function that differentially weights cloud events
relative to the clear-sky envelope (e.g., scaling the per-sample MSE by the
departure of observed GHI from the clear-sky reference), or dedicated
cloud-event detection as a preprocessing step.

\subsection{Practical implications}

\begin{itemize}
  \item For \textbf{El~Paso-type sites} (high irradiance, moderate variability),
        GraphSAGE-LSTM with Optuna-tuned hyperparameters is the recommended model
        at 6h, reducing RMSE over persistence by approximately 63.6\%. At 3h, the
        two architectures achieve statistically indistinguishable mean skill
        (both $\skillday \approx 0.42$), but GraphSAGE-LSTM's substantially
        lower seed-to-seed variance ($\pm 0.039$ vs.\ $\pm 0.104$ for
        ResNet-LSTM) makes it the more reliable choice for production
        deployment at that horizon. At 1h, the system should fall back to
        persistence or a light-weight nowcasting model.

  \item For \textbf{Uniandes-type sites} (tropical mountain, high convective
        variability), even the best models achieve markedly lower skill than at
        El~Paso ($\skillday \leq 0.42$ vs.\ $0.64$ at 6h); operators should
        calibrate reserve margins to the local persistence baseline rather than
        assume that skill levels transfer across climates. GraphSAGE-LSTM remains
        the most consistent top performer at 3h and 6h, but gains over
        persistence are smaller and more sensitive to careful hyperparameter
        tuning.
\end{itemize}

Within a site, the best architecture also depends on the horizon: persistence
is hard to beat at 1h in the clear-sky-dominated El~Paso regime, while
GraphSAGE-LSTM is the most consistent top performer at 3h and 6h at both
sites. A horizon-aware portfolio strategy --- a light-weight nowcasting
fallback at very short horizons, spatial deep learning at multi-hour horizons
--- could optimise the accuracy-cost tradeoff for operational deployment.

\subsection{Why epistemic uncertainty matters on its own}
\label{ssec:uq_discussion}

We argue that grid operators are better served by probabilistic outputs ---
prediction intervals that are well-calibrated across conditions of varying
cloud cover --- than by point estimates alone. Beyond calibration, however,
the central premise of this work is that knowing \emph{whether the model
itself is confident} in a forecast is operationally valuable, independent of
the width of any interval.

\textbf{Why interval width alone is not enough.} A prediction interval's
width alone does not tell an operator whether the underlying model was well
supported by its training data when producing a given forecast. Two
forecasts can look equally plausible on their face while differing in this
respect: one produced under a satellite-imagery pattern the model has seen
extensively, the other under a rare or under-represented pattern. An
operator cannot tell these apart from a point forecast alone, yet the
appropriate response differs --- trust the forecast under normal deployment
in the first case; flag it as low-confidence and prefer a conservative
fallback (e.g.\ persistence, or a wider manually-set margin) in the second.
We do not attempt to separately model the irreducible aleatoric
stochasticity of cloud formation here (Section~\ref{sec:discussion},
Limitations); the SGLD ensemble targets specifically the reducible,
data-dependent share of uncertainty.

\textbf{Epistemic uncertainty as a data-quality and deployment-readiness
signal.} Because $\sigma_\text{epi}$ (Eq.~\ref{eq:epistemic}) reflects
disagreement among SGLD posterior samples, it is expected to be larger for
satellite-imagery patterns under-represented in the training period (e.g.\
rare cloud morphologies, seasons with sparse data) and to shrink as the
training set grows. This makes $\sigma_\text{epi}$ a natural diagnostic for
\emph{when} a forecasting pipeline should be retrained or supplemented with
additional data at a given site, complementing the purely architecture-level
comparison of Sections~\ref{ssec:resnet}--\ref{ssec:mlp} and~\ref{ssec:ablation}.

\subsection{Implications for solar-integrated power system operation}

The results presented here connect directly to operational decision-making in
solar-integrated power systems.

\textbf{Situational awareness under cloud-induced uncertainty.}
By learning spatially-aware representations of incoming cloud systems from
satellite imagery, GraphSAGE-LSTM may support earlier detection of impending
irradiance ramps --- the primary source of generation-side uncertainty in
PV-heavy grids. The 63.6\% RMSE reduction over persistence at 6h for
El~Paso translates to significantly tighter dispatch intervals and reduced
reserve requirements.

\textbf{Epistemic-aware dispatch.}
The SGLD-based epistemic estimate (Section~\ref{ssec:uq}) converts any point
forecast into both an interval and a signal of how well-supported it is by
the model's training data. For a system operator, this supports two
distinct, complementary actions tied to the same forecast: (i) sizing
reserve margins to the interval $\bar\mu \pm z_{1-\alpha/2}\sigma_\text{epi}$
for routine risk management; and (ii) flagging forecasts with high
$\sigma_\text{epi}$ as candidates for a more conservative fallback strategy
or for prioritised inclusion in future retraining data --- a distinction
unavailable from a point forecast alone.

\subsection{Limitations}

\textbf{Geographic scope.} Both sites are located in Colombia and represent
two tropical climate sub-types. Generalisation to other tropical or subtropical
regions requires additional validation.

\textbf{FlatMLP ablation in progress.} The FlatMLP experiments are currently
completing their Optuna runs (10 of the 24 site $\times$ horizon $\times$ seed
configurations evaluated at the time of writing). Final results will be
incorporated to quantify what fraction of deep-learning skill arises from
spatial encoding versus temporal dynamics alone.

\textbf{FlatMLP is not a pure spatial ablation.} Because FlatMLP also
replaces the shared LSTM with a feed-forward network over the concatenated
sequence (Section~\ref{ssec:mlp}), its skill gap relative to ResNet-LSTM and
GraphSAGE-LSTM reflects the joint loss of spatial structure \emph{and} of
the recurrent temporal model, and should be read as an approximate upper
bound on the spatial-encoding contribution rather than a clean, single-factor
estimate.

\textbf{Graph-construction ablation uses fewer seeds.} The weighted $k$-NN
GraphSAGE-LSTM variant (Table~\ref{tab:ablation}) is evaluated with two
random seeds rather than four, for the compute-budget reasons described in
Section~\ref{ssec:optuna}. This narrower seed budget should be kept in mind
when comparing its mean skill against the four-seed fixed-graph baseline;
the comparison should be treated as indicative pending a larger seed budget.

\textbf{Epistemic uncertainty has not yet been empirically validated across
all backbones.} The SGLD sampling pipeline (Section~\ref{ssec:uq}) is
implemented but has not yet been run to completion across all site-horizon
backbones (Table~\ref{tab:decomposition}). The claim that $\sigma_\text{epi}$
tracks data scarcity (Section~\ref{ssec:uq_discussion}) is currently a
hypothesis motivating the design, not yet an empirical finding; it will be
assessed quantitatively once sampling completes.

\textbf{No separate aleatoric uncertainty estimate yet integrated.} The
results reported here quantify only the epistemic (reducible,
data-dependent) share of predictive uncertainty via the SGLD ensemble;
consequently, the reported interval $\bar\mu \pm z_{1-\alpha/2}\sigma_\text{epi}$
should be read as a lower bound on total predictive uncertainty rather than
a fully calibrated interval. An aleatoric variance network
(Section~\ref{ssec:variance}), following \citet{yi2025cooperative}'s
cooperative strategy, is being trained separately across the available
backbones, but its output is not yet integrated into the SGLD sampler or
the reported interval; completing this integration to obtain a full
aleatoric/epistemic decomposition is ongoing work (Section~\ref{sec:conclusion}).

\textbf{Single-step output.} Current models produce a single-step forecast at
horizon $H$. Simultaneous multi-step output (forecasts for all three horizons
from a single forward pass) could improve consistency and reduce inference cost
in real-time operational forecasting pipelines.
